Safe Transferable Regions

### 2015-11-20 paper/type-system.tex
Merged paper to techrep.

@@

\input{fb-staticsem}

@@

% Like methods, functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
\FB extends FGJ's expression language with a lambda expression and an
application expression ($e\inang{\ralloc\rbar}(\bar{e})$) to define
and apply functions\footnote{We distinguish between functions and
methods. The former result from lambda expressions, whereas the latter
come from class definitions}. Functions, like methods, exhibit
constrained region polymorphism, as evident in their arrow region type
($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
\tau$).
% at the prenex position, similar to ML type schemes. However,
% unlike ML, we don't distinguish between types and type schemes; any of
% the $\tau$'s in the arrow type can themselves be region-parametric
% arrow types. Like with methods, region parameters on functions can be
% instantiated when they are applied.
% In this respect, our region type system is more like
% System F's type system, which admits higher-rank parametric
% polymorphism. Like System F, \FB provides a region instantiation
% expression ($e\inang{\ralloc\rbar}$), and a region generalization
% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
% instantiate and generalize region variables. 
Since a function closure can escape the context in which it is
created, it is important to keep track of the region in which it is
created in order to avoid unsafe dereferences. The $\rgn$ annotation
above the arrow in the arrow type denotes the allocation region of the
corresponding closure. 
% Note that it is important to distinguish
% between the allocation context argument ($\rhoalloc$) of a function
% and the allocation region ($\rgn$) of its closure. In \name, the later
% corresponds to the region where a \C{Func} object is allocated, while
% the former corresponds to the region where it is applied
. 
% For instance, in the following example:
% \begin{codejava}
% letregion $\rgn$ {
%   let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
%   in f
% }
% \end{codejava}
% The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
% \ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
% in $\rgn$, and (b). when executed under an allocation context
% $\rho^a$, the closure returns an object allocated in $\rho^a$.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{The full formal development can be found in the
supplement}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable
($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
was declared to extend.}, and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
$\A$ also contains $\rhomap$, 
% a map ($\rhomap \in \rgn \mapsto
% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
% transferable regions to their current state in the FSM from
% Fig.~\ref{fig:region-fsm}.
which is primarily an artifact to facilitate the type safety proof,
and can be ignored while type checking user-written programs
in \FB. The context for the expression typing judgment also includes:
\begin{itemize}
\item A type environment ($\env \in x \mapsto \tau$) that contains the
type bindings for variables in scope, and 
\item The static identifier ($\rgn^a$) of the allocation context for
the expression is being typechecked.  
\end{itemize}
Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
class table ($CT \in cn \mapsto D$) that maps class names to their
definitions in \FB.
% \begin{figure}

% \begin{codejava}
% class Region<a $\extends$ Object><${\rhoalloc}$> $\extends$ Object<$\rhoalloc$>{
%   <${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk;
%   Region(<${\rho}$>unit$\xrightarrow{\rhoalloc}$a@$\rho$ thunk) {
%     super(); 
%     this.thunk = thunk;
%   }
%   unit transfer$\inang{\rho}$(unit u) {
%     return u;
%   }
% }
% \end{codejava}

% \caption{The stub used for $\RgnZ$ class}
% \label{fig:region-stub}
% \end{figure}

\noindent The well-formedness judgment on region types
($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
judgments on core types \footnote{We use a double-piped turnstile
($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
($\vdash$) for those in \FB.}. The class table ($\absof{CT}$) for such
judgments is derived from \FB's class table ($CT$) by erasing all
region annotations on types, and region arguments in expressions
($\absof{\cdot}$ denotes the region erasure operation). 
% For judgments in FGJ, we use a
% class table that maps class names to their region-erased definitions.
% We denote the region-erased class table as ($\absof{CT}$). The precise
% semantics of region erasure ($\absof{\cdot}$) are defined
% in~\cite{techrep}, but it can be understood as an operation that
% erases region annotations on types. The class table is assumed to bind
% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
% This mostly eliminates the need to treat $\RgnZ$ objects, which double
% up as transferable region handlers, as a special case in semantics.
The well-formedness rule for class types
($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
the safety property that prevents objects from containing unsafe
references. It does so by insisting that regions $\ralloc\rbar$
satisfy the constraints ($\phi$) imposed by the class on its region
parameters. The later is enforced by checking the validity of $\phi$,
with actual region arguments substituted\footnote{The notation
$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
region parameters, under the conditions ($\A.\phicx$) guaranteed by
the context. The semantics of this sequent is straightforward, and
follows directly from the properties of outlives and equality
relations. For any well-formed core type $T$, $T@\rgn$ is a
well-formed region type if $\rgn$ is a valid region. The type
$\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
$\toprgn$ is a special immortal region that outlives every other live
region. This arrangement allows \C{\RgnZ} handlers to be aliased and
referenced freely from objects in various regions, regardless of their
lifetimes. On the flip side, this also opens up the possibility of
references between transferable regions, which become unsafe in
context of the recipient's address space. Fortunately, such
references are explicitly prohibited by the type rule of \C{\RgnZ}
objects, as described below.

@@

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
order to prevent the new region from incorrectly assuming any existing
outlives relationships on an eponymous region. Provided this is
satisfied, the expression ($e$) under \C{letregion} is then
typechecked assuming that the new region is live ($\rgn \in \rhoenv$)
and that it is outlived by all existing live regions ($\rhoenv
\outlives \rgn$). The result of a \C{letregion} expression must have a
type that is well-formed under a context not containing the new
region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region\footnote{The supplement contains a formal
proof of this claim}.

@@

The supplement defines a small-step operational semantics for \fbname
via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state. The supplement
defines a type rule for such values, allowing them to be typed.
Type safety theorem is now stated thus:
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-type-safety}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. Notably, this result could not have been
established if it were possible for a transferable region to contain
references to objects outside the region.

>>>>

\input{fb-auxdef}

@@

\input{fb-staticsem}
\input{fb-morewfrules}

@@

% Like methods, functions can also be region-polymorphic. 
% A lambda expression defines a region-polymorphic
% multi-argument function closure parameterized over function's
% allocation context parameter. 
\FB extends FGJ's expression language with a lambda expression and an
application expression ($e\inang{\ralloc\rbar}(\bar{e})$) to define
and apply functions\footnote{We distinguish between functions and
methods. The former result from lambda expressions, whereas the latter
come from class definitions}. Functions, like methods, exhibit
constrained region polymorphism, as evident in their arrow region type
($\inang{\rhoalloc\rhobar \,|\, \phi}\bar{\tau} \xrightarrow{\rgn}
\tau$). Note that any of the $\tau$'s in the arrow type can themselves
be region-parametric arrow types. In this respect, our region type
system is comparable to System F's type system, which admits
higher-rank parametric polymorphism. 
% at the prenex position, similar to ML type schemes. However,
% unlike ML, we don't distinguish between types and type schemes; any of
% the $\tau$'s in the arrow type can themselves be region-parametric
% arrow types. Like with methods, region parameters on functions can be
% instantiated when they are applied.
% In this respect, our region type system is more like
% System F's type system, which admits higher-rank parametric
% polymorphism. Like System F, \FB provides a region instantiation
% expression ($e\inang{\ralloc\rbar}$), and a region generalization
% expression ($\Lambdaexp{\rhoalloc\rhobar \,|\, \phi}{e}$) to
% instantiate and generalize region variables. 
Note that a lambda expression creates a closure, which can escape the
context in which it is created. It is therefore important to keep track of
the region in which a closure is allocated in order to avoid unsafe
dereferences. The $\rgn$ annotation above the arrow in the arrow
type denotes the allocation region of the corresponding closure. Note
that it is important to distinguish between the allocation context
argument ($\rhoalloc$) of a function and the allocation region
($\rgn$) of its closure. In \name, the later corresponds to the region where
a \C{Func} object is allocated, while the former corresponds to the
region where it is applied.  For instance, in the following example:
\begin{codejava}
letregion $\rgn$ {
  let f = $\lambda{\inang{\rho^a}}$().$\,$new Object$\inang{\rho^a}$() 
  in f
}
\end{codejava}
The type of \C{f} is $\inang{\rho^a}\unitZ \xrightarrow{\rgn}
\ObjZ\inang{\rho^a}$, coveying that (a). \C{f}'s closure is allocated
in $\rgn$, and (b). when executed under an allocation context
$\rho^a$, the closure returns an object allocated in $\rho^a$.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains the entire set of the rules. The
rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of static identifiers of regions
that are estimated to be live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their bounds\footnote{A bound of a type variable
($\tyvar$) in FGJ~\cite{fgj} is the class ($\fgjN$) that type variable
was declared to extend.}, and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
$\A$ also contains $\rhomap$, 
% a map ($\rhomap \in \rgn \mapsto
% \{\CLOSED,\OPEN,\XFERRED\}$) from unique region identifiers of
% transferable regions to their current state in the FSM from
% Fig.~\ref{fig:region-fsm}.
which is primarily an artifact to facilitate the type safety proof,
and can be ignored while type checking user-written programs
in \FB. The context for the expression typing judgment also includes:
\begin{itemize}
\item A type environment ($\env \in x \mapsto \tau$) that contains the
type bindings for variables in scope, and 
\item The static identifier ($\rgn^a$) of the allocation context for
the expression is being typechecked.  
\end{itemize}
Like the judgments in FGJ~\cite{fgj}, all the judgments defined by the
rules in Fig.~\ref{fig:fb-staticsem} are implicitly parameterized on a
class table ($CT \in cn \mapsto D$) that maps class names to their
definitions in \FB.

\noindent The well-formedness judgment on region types
($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
judgments on core types \footnote{We use a double-piped turnstile
($\Vdash$) for judgments in FGJ~\cite{fgj}, and a simple turnstile
($\vdash$) for those in \FB.}. The class table ($\absof{CT}$) for such
judgments is derived from \FB's class table ($CT$) by erasing all
region annotations on types, and region arguments in expressions
($\absof{\cdot}$ denotes the region erasure operation).  The
well-formedness rule for class types
($B\inang{\tbar}\inang{\ralloc\rbar}$) is responsible for enforcing
the safety property that prevents objects from containing unsafe
references. It does so by insisting that regions $\ralloc\rbar$
satisfy the constraints ($\phi$) imposed by the class on its region
parameters. The later is enforced by checking the validity of $\phi$,
with actual region arguments substituted\footnote{The notation
$[a/b](e)$ stands for "$a$ is substituted for $b$ in $e$"} for formal
region parameters, under the conditions ($\A.\phicx$) guaranteed by
the context. The semantics of this sequent is straightforward, and
follows directly from the properties of outlives and equality
relations. For any well-formed core type $T$, $T@\rgn$ is a
well-formed region type if $\rgn$ is a valid region. The type
$\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
$\toprgn$ is a special immortal region that outlives every other live
region. This arrangement allows \C{\RgnZ} handlers to be aliased and
referenced freely from objects in various regions, regardless of their
lifetimes. On the flip side, this also opens up the possibility of
references between transferable regions, which become unsafe in
context of the recipient's address space. Fortunately, such references
are explicitly prohibited by the type rule of \C{\RgnZ}
objects, as described below.

@@

\input{fb-opsem}

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\rgn \notin \rhoenv$). This condition is needed in
order to prevent the new region from incorrectly assuming any existing
outlives relationships on an eponymous region. Provided this is
satisfied, the expression ($e$) under \C{letregion} is then
typechecked assuming that the new region is live ($\rgn \in \rhoenv$)
and that it is outlived by all existing live regions ($\rhoenv
\outlives \rgn$). The result of a \C{letregion} expression must have a
type that is well-formed under a context not containing the new
region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region\footnote{The appendix contains a formal
proof of this claim}.

@@

Fig.~\ref{fig:fb-opsem} defines a small-step operational semantics for
\fbname via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

To help state the type safety theorem, we define the syntactic class of values:
\begin{smathpar}
\begin{array}{lclcl}
v & \in & \mathtt{values} & \coloneqq & \C{new}\; \fbN(\bar{v}) \ALT
\lambdaexp{\ralloc}{\rhoalloc\rhobar\,|\,\phi}{\taubar \; \xbar}{e}\\
  &     & & & \C{new}\; \RgnZT{\rgn}(v)\\
\end{array}
\end{smathpar}
Note that for $\C{new}\; \RgnZT{\rgn}(v)$ to be considered a value,
$\rgn \neq \toprgn$. The semantics reduces a $\C{new}\;
\RgnZT{\toprgn}(e)$ expression in the user program to a runtime
$\C{new}\; \RgnZT{\rgn}(v)$ value that is tagged with a unique
identifier ($\rgn$) for this region. A binding is also added to
$\rhomap$, mapping $\rgn$ to the ``closed'' state.
Fig.~\ref{fig:fb-staticsem} defines a type rule for such values,
allowing them to be typed.  Type safety theorem is now stated
thus\footnote{Theorems~\ref{thm:fb-type-safety-stmt}
and~\ref{thm:fb-transfer-safety-stmt} are restated and proved in the
appendix}:
\begin{theorem}
\emph{(\textbf{Type Safety})}
\label{thm:fb-type-safety-stmt}
$\forall e, \tau, \rhoenv, \rhomap, \rgn$, such that $\rgn \in
\rhoenv$ and $\tywf{\rhoenv}{\phicx}$, if $\hastyp{\emptyA,\rgn,
\cdot}{e}{\tau}$, then either $e$ is a value, or $e$ raises an
exception ($\redstoo{(e,\rhomap)}{\invalidexn}$), or there exists an
$e'$ and a $\rhomap'$ such that $\redstoo{(e,\rhomap)}{(e',\rhomap')}$
and $\hastyp{\emptyASigp,\rgn, \cdot}{e}{\tau}$.
\end{theorem}

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-transfer-safety-stmt}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. Notably, this result could not have been
established if it were possible for a transferable region to contain
references to objects outside the region.
